Beyond May: Complexity-stability relationships \\ in disordered dynamical systems

### 2023-05-07 papers/equilibrium.tex

@@

Equipped with the equilibrium distribution $P(x^*)$, we can compute the maximal heterogeneity $\sigma_c$, for a given $N$ and $\mu$, for which the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds. 
A convenient way to test the complexity-stability properties of the system, consists in studying how this condition compart the $(\sigma \sqrt{N},\mu N)$ plan, conventionally used to summarize the properties of large random dynamical systems~\cite{bunin2017ecological}, in stable and unstable parameter regions. 
Everything else fixed, an increase in the number of degrees of freedom $N$ will move the system along a square-root trajectory in that plane and we can test if this has stabilizing effects on the system or if the opposite is true.
In Fig.~\ref{fig: stability line + sims} the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ is portrayed in the $(\sigma \sqrt{N},\mu N)$ plane, compared with results from simulations (where stability is defined as stable, full, coexistence of the system).
An increase in the number of degrees of freedom $N$ will eventually bring the system in the stable region, or preserve stability if already stable. 
In other words, we are in the ``anti-May" phase.
(For comparison the stability boundary for GLV is a horizontal line in the $(\sigma \sqrt{N},\mu N)$ plane, defined by $\sigma_c\sqrt{N} = \mu_s$~\cite{bunin2017ecological}.)

We conclude by testing the robustness of our findings in the case of heterogeneous interactions in the paradigmatic case of power laws. Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$, defined as the value above which full stable coexistence is less probable than $50\%$.
(In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model.) This plot illustrates the transition between a ``May" phase for $\alpha \geq \beta$ (where stability decreases with $N$, $\mu$, and $\sigma$) and an ``anti-May" phase for $\alpha < \beta$, even in the presence of heterogeneous interactions.

>>>>

Equipped with the equilibrium distribution $P(x^*)$, we can compute, for given $N$ and $\mu$, the maximal heterogeneity $\sigma_c$ consistent with the linear stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ holds.

@@

The properties of large random dynamical systems is often portrayed in often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}. Everything else fixed, an increase of the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane. In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.

In Fig.~\ref{fig: stability line + sims} we plot the stability condition $N\langle (\sigma/\psi(x^*))^2\rangle < 1$ for $\alpha = 1$, $\beta = 3/2$ derived from \eqref{eq: dist general} (solid line) together with results from simulations (shading). Here, because the boundary between the stable and unstable phases is a straight line with positive slope, increasing $N$ will eventually bring the system in the stable region. This behavior corresponds to the ``anti-May" phase defined previously.

Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$. (Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations we use $\mu_s = \mu$, which would never be stable in the GLV model. Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

### 2023-05-07 papers/stability.tex

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = - g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in~\eqref{eq: psi}. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}
Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, the dependence on equilibrium values $x_i^*$ (which, in turn, can depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

>>>>

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{equation}
    M_i = - g(x_i^*)h'(x_i^*)\psi(x_i^*),\quad L = g(x_i^*), \quad R = h'(x_i^*),
\end{equation}
where $\psi(x)$ is the function defined in~\eqref{eq: psi}. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{figs/cavity.pdf}
    \caption{The equilibrium distribution with power-law self- and cross-interactions is accurately reproduced by 
    Eq.~\eqref{eq: dist general} and its Gaussian approximation.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$, $\mu=\mu_s=10^{-2}$, $\sigma=5\cdot 10^{-3}$.}
    \label{fig: cavity sol.}
\end{figure}
Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < (\mu_s-\mu)$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.
